\section{Introduction/Setup}
%
As with any programming language, picking up an ITP system first requires
installing the system.
In this guide we use the HOL4~\cite{HOL4web} theorem prover.
The guide also uses a tool called \emph{Lassie}, equipping HOL4 with training
wheels.

To follow this guide, \texttt{polyML}, \texttt{Apache Ant}, and \texttt{ruby}
must be installed.
For \texttt{polyML}, we recommend downloading the latest version from git
(\url{https://github.com/polyml/polyml}).
Both \texttt{Apache Ant}, and \texttt{ruby} can be installed through the system
package manager.
We provide a setup script that makes sure all the dependencies are installed,
and downloads and installs the latest versions of HOL4 and Lassie with this
guide (\url{https://people.mpi-sws.org/~hbecker/setupHOL4.sh}).
Running the setup script temporarily sets the variables \texttt{HOLDIR}, and
\texttt{LASSIEDIR} to the directories where HOL4 and Lassie have been installed.
To use both tools later, these variables must be be configured permanently
before running HOL4.

\subsection{Running HOL4}

We will explain how to interact using HOL4 using the emacs editor.
There also exist plugins for the vim, and Sublime text editors, and the overall
interaction is the same, only the keybindings may change.

To interact with HOL4 within emacs, it suffices to append the following lines to
the emacs configuration file:
\begin{lstlisting}
(load-file "<HOL install dir>/tools/hol-unicode.el")
(load-file "<HOL install dir>/tools/hol-input.el")
(load-file "<HOL install dir>/tools/holscript-mode.el")
(load-file "<HOL install dir>/tools/hol-mode.el")
\end{lstlisting}

After loading the files, respectively restarting emacs, HOL4 files will by
default open in the HOL4 mode.

In general, HOL4 files always end with the suffix \texttt{Script.sml}.
For example, a theory about natural numbers is called
\texttt{naturalNumbersScript.sml}
In HOL4 speak, files ending with \texttt{Script.sml} are called \emph{script files}.

To interact with HOL4 we must first star ta read-eval-print loop (REPL), similar
to interpreter of programming languages like ruby or python.
After opening a script file in emacs, HOL4 is started by presssing
\ekey{M-h H}.
emacs will then interactively prompt for a position of the HOL4 REPL.